---
title: "R Notebook"
output: html_notebook
---


```{r}
library(tidyverse)
```

```{r}
loan <- read_csv("~/Documents/Data Science 2022/Mid Course Project/Data/state_TN_actions_taken.csv")
```

The analysis focuses on loan applications for the purpose of 'Home Purchase' (loan_purpose == 1), 'Not primarily for a business or commercial purpose' (business_or_commercial_purpose == 2), Occupancy type for the dwelling is "1 - Principal residence", and all loan type except "Veterans Affairs guaranteed (VA)" (loan_type!=3)

```{r}
mortgage_loan <- loan %>%
  filter(loan_purpose == 1, 
         business_or_commercial_purpose == 2,
         occupancy_type == 1,
         loan_type != 3, applicant_age != "8888") %>%
  mutate(loan_app_denied = ifelse(action_taken==3,1,0)) %>%
  mutate(debt_to_income_group = case_when(
    debt_to_income_ratio == "30%-<36%" ~ "30%-<40%",
    as.numeric(debt_to_income_ratio) %in% 36:39 ~ "30%-<40%",
    as.numeric(debt_to_income_ratio) %in% 40:49 ~ "40%-<50%",
    TRUE ~ debt_to_income_ratio
    ))
```


```{r}
mortgage_loan_summary <- 
  mortgage_loan %>% 
  group_by(derived_race) %>%
  summarize(application_count = n(), denial_count = sum(loan_app_denied)) %>%
  mutate(pct_denial = 100*denial_count/application_count) %>%
  filter(derived_race != c("American Indian or Alaska Native","Native Hawaiian or Other Pacific Islander"))

mortgage_loan_summary_allraces <- 
  mortgage_loan %>% 
  group_by(derived_race) %>%
  summarize(application_count = n(), denial_count = sum(loan_app_denied)) %>%
  mutate(pct_denial = 100*denial_count/application_count)

mortgage_loan_debt_summary <-
  mortgage_loan %>% 
  group_by(derived_race, debt_to_income_group) %>%
  summarize(application_count = n()) %>%
  filter(derived_race != "American Indian or Alaska Native" & derived_race != "Native Hawaiian or Other Pacific Islander")
```

For the purpose of this analysis, the following plot ignores "American Indian or Alaska Native" and "Native Hawaiian or Other Pacific Islander" because they make up less than 0.01% of total mortgage applications. 

```{r}
mortgage_loan_summary %>% 
  ggplot(aes(x=derived_race, y=pct_denial, fill=derived_race)) +
  geom_col() +
  theme(axis.text.x = element_text(angle = 60, hjust = 1)) +
  scale_x_discrete(name = "Race") +
  scale_y_continuous(name = "% Denial") 
```

```{r}
mortgage_loan_debt_summary %>% 
  ggplot(aes(y=application_count, x=debt_to_income_group, fill=derived_race)) +
  geom_col(position="dodge") +
  scale_y_continuous(trans = "log10") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  xlab("Debt-to-Income (DTI) Groups") +
  ylab("Total Mortgage Applications") +
  ggtitle("DTI by Race")

mortgage_loan_debt_summary %>% 
  ggplot(aes(y=application_count, x=debt_to_income_group, fill=derived_race)) +
  geom_col(position="dodge") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) 
```
1 - Debt-to-income ratio
2 - Employment history
3 - Credit history
4 - Collateral
5 - Insufficient cash (downpayment, closing costs)
6 - Unverifiable information
7 - Credit application incomplete
8 - Mortgage insurance denied
9 - Other
10 - Not applicable

```{r}
denial_reasons <- 
  data.frame(denial_reason_code=1:10,
             denial_reason_text=c("Debt-to-income ratio",
                                  "Employment history",
                                  "Credit history",
                                  "Collateral",
                                  "Insufficient cash\n (downpayment, closing costs)",
                                  "Unverifiable information",
                                  "Credit application\n incomplete",
                                  "Mortgage insurance denied",
                                  "Other",
                                  "NA"))
total_race_denial_count <- 
  mortgage_loan %>%
  filter(loan_app_denied==1) %>%
  group_by(derived_race) %>%
  summarize(total_denials_by_race = n())

total_age_denial_count <- 
  mortgage_loan %>%
  filter(loan_app_denied==1) %>%
  group_by(applicant_age) %>%
  summarize(total_denials_by_age = n())

total_sex_denial_count <- 
  mortgage_loan %>%
  filter(loan_app_denied==1) %>%
  group_by(derived_sex) %>%
  summarize(total_denials_by_sex = n())


denial_reasons_count_by_race <-
  mortgage_loan %>%
  select(loan_app_denied, derived_race, `denial_reason-1`, `denial_reason-2`, `denial_reason-3`, `denial_reason-4`) %>%
  pivot_longer(cols = c(`denial_reason-1`, `denial_reason-2`, `denial_reason-3`, `denial_reason-4`),
               values_to = "denial_reason_code") %>%
  filter(is.na(denial_reason_code)==F, denial_reason_code!=1111, 
         loan_app_denied == 1, derived_race %in% c("White","Asian","Black or African American")) %>%
  group_by(derived_race, denial_reason_code) %>%
  summarize(denial_reason_count = n()) %>%
  left_join(denial_reasons, by = "denial_reason_code") %>%
  left_join(total_race_denial_count, by = "derived_race") %>%
  mutate(pct_race_denial = round(100*(denial_reason_count/total_denials_by_race),2))

denial_reasons_count_by_age <-
  mortgage_loan %>%
  select(loan_app_denied, applicant_age, `denial_reason-1`, `denial_reason-2`, `denial_reason-3`, `denial_reason-4`) %>%
  pivot_longer(cols = c(`denial_reason-1`, `denial_reason-2`, `denial_reason-3`, `denial_reason-4`),
               values_to = "denial_reason_code") %>%
  filter(is.na(denial_reason_code)==F, denial_reason_code!=1111, loan_app_denied == 1, applicant_age != ">74") %>%
  group_by(applicant_age, denial_reason_code) %>%
  summarize(denial_reason_count = n()) %>%
  left_join(denial_reasons, by = "denial_reason_code") %>%
  left_join(total_age_denial_count, by = "applicant_age") %>%
  mutate(pct_age_denial = round(100*(denial_reason_count/total_denials_by_age),2))

denial_reasons_count_by_sex <-
  mortgage_loan %>%
  select(loan_app_denied, derived_sex, `denial_reason-1`, `denial_reason-2`, `denial_reason-3`, `denial_reason-4`) %>%
  pivot_longer(cols = c(`denial_reason-1`, `denial_reason-2`, `denial_reason-3`, `denial_reason-4`),
               values_to = "denial_reason_code") %>%
  filter(is.na(denial_reason_code)==F, denial_reason_code!=1111, loan_app_denied == 1, derived_sex != "Sex Not Available") %>%
  group_by(derived_sex, denial_reason_code) %>%
  summarize(denial_reason_count = n()) %>%
  left_join(denial_reasons, by = "denial_reason_code") %>%
  left_join(total_sex_denial_count, by = "derived_sex") %>%
  mutate(pct_sex_denial = round(100*(denial_reason_count/total_denials_by_sex),2))

save(mortgage_loan_denial_reasons, file = '/Users/thidathornvanitsthian/Documents/Data Science 2022/loan-denial-rate-TN/Data/mortgage_loan_denial_reasons')
```
Facet Wrap ggplot % Denial by Race, Age, Sex

```{r}
plot_race <- ggplot(denial_reasons_count_by_race) +
    facet_wrap(~derived_race, ncol = 1) +
    geom_col(aes(x=factor(denial_reason_text), y=pct_race_denial, fill=factor(derived_race)), show.legend = F) +
    theme(axis.text.x = element_text(angle=60,hjust=1)) +
    scale_x_discrete(name = "Denial Reasons") +
    scale_y_continuous(name = "% Denial by Race") 
  
save(plot_race, file = '/Users/thidathornvanitsthian/Documents/Data Science 2022/loan-denial-rate-TN/Data/plot_race')

# plot_age <- ggplot(denial_reasons_count_by_age) +
#     facet_wrap(~applicant_age, ncol = 1) +
#     geom_col(aes(x=factor(denial_reason_text), y=pct_age_denial, fill=factor(applicant_age)), show.legend = F) +
#     theme(axis.text.x = element_text(angle=60,hjust=1)) +
#     scale_x_discrete(name = "Denial Reasons") +
#     scale_y_continuous(name = "% Denial by Age Group") 

plot_age <- ggplot(denial_reasons_count_by_age) +
    geom_col(aes(fill=factor(denial_reason_text), y=pct_age_denial,x=factor(applicant_age)), show.legend = T) +
    theme(axis.text.x = element_text(angle=60,hjust=1)) +
    xlab("Age Groups") +
    ylab("% Denial Reasons") +
    scale_fill_discrete(name = "Age\nGroups")

save(plot_age, file = '/Users/thidathornvanitsthian/Documents/Data Science 2022/loan-denial-rate-TN/Data/plot_age')

plot_sex <- ggplot(denial_reasons_count_by_sex) +
    facet_wrap(~derived_sex, ncol = 1) +
    geom_col(aes(x=factor(denial_reason_text), y=pct_sex_denial, fill=factor(derived_sex)), show.legend = F) +
    theme(axis.text.x = element_text(angle=60,hjust=1)) +
    scale_x_discrete(name = "Denial Reasons") +
    scale_y_continuous(name = "% Denial by Sex") 

save(plot_sex, file = '/Users/thidathornvanitsthian/Documents/Data Science 2022/loan-denial-rate-TN/Data/plot_sex')

```
Facet Wrap ggplot Denial Count By Race, Age, Sex

```{r}
ggplot(mortgage_loan_denial_reasons) +
    facet_wrap(~derived_race, ncol = 1) +
    #geom_bar(aes(x=factor(denial_reason_text), fill=factor(derived_race)), position="dodge") +
    geom_bar(aes(x=factor(denial_reason_text), fill=factor(derived_race)), show.legend = F) +
    theme(axis.text.x = element_text(angle=60,hjust=1)) +
    scale_x_discrete(name = "Denial Reasons") +
    scale_y_continuous(name = "Denial Frequency") 
    #scale_fill_discrete(name = "Race")
```

```{r}
mortgage_loan$num_reasons <- apply(mortgage_loan[c("denial_reason-1", "denial_reason-2", "denial_reason-3", "denial_reason-4")], 1,function(x){length(grep(F,is.na(x)))})
```

Logit Model: Age, DTI, Income, Race, Sex

```{r}
#library(lme4)

mortgage_loan_glm <- 
  mortgage_loan %>%
  filter(debt_to_income_group != "Exempt" & !is.na(debt_to_income_group),
         income > 0 & income < 200,
         loan_amount < 500000)

# To control for fixed and random effect, use lme4

# glm_ft <- 
#   glmer(loan_app_denied ~ 
#         (1|applicant_age) + (1|debt_to_income_group) + income + (1|derived_race) + 
#         (1|derived_sex), 
#       family = binomial(link = "logit"),
#       data = mortgage_loan_glm)
# summary(glm_ft)
# coef(glm_ft)
# ranef(glm_ft)


glm_ft <-
  glm(loan_app_denied ~
        factor(applicant_age) + factor(debt_to_income_group) + income + factor(derived_race) +
        factor(derived_sex) -1,
      family = binomial(link = "logit"),
      data = mortgage_loan_glm)
summary(glm_ft)


save(glm_ft, file = '/Users/thidathornvanitsthian/Documents/Data Science 2022/loan-denial-rate-TN/Data/glm_ft')

mortgage_loan_glm$glm_ft_predict <- 
  ifelse(predict(glm_ft, mortgage_loan_glm, type = "response") < 0.5, 0, 1)


#Confusion Matrix
table(mortgage_loan_glm[c("loan_app_denied","glm_ft_predict")])
#https://www.statology.org/confusion-matrix-in-r/

100*table(mortgage_loan_glm$glm_ft_predict==mortgage_loan_glm$loan_app_denied)/nrow(mortgage_loan_glm)
```


Calibration Curve Data
```{r}
mortgage_loan_glm$glm_ft_predict_probability <- 
  predict(glm_ft, mortgage_loan_glm, type = "response") 

bins <- seq(from = 0, to = 1, by=0.1)
mortgage_loan_glm %>%
  mutate(group=cut(glm_ft_predict_probability, bins)) %>%
  group_by(group) %>%
  summarize(mean_group = mean(loan_app_denied), count_group = n()) %>%
  View()
```


Test Plot
```{r}
a <- 1-glm_ft_prediction$fit[1]
b <- glm_ft_prediction$se.fit[1]

x <- seq(0,1,length=100)
height <- dnorm(x,a,2*b)

plot_df <- data.frame(x=x,height=height)

ggplot(plot_df, aes(x=x, y=height)) +
  geom_line() +
  scale_y_continuous(name = "", labels = NULL) +
  ggtitle(paste0("You are estimated to have between a ", 100*round(1-(a+b*1.96), digits = 2), 
      " and a ", 100*round(1-(x-b*1.96), digits = 2),
      "% probability of being approved\n a mortgage loan in TN"))
```


Create CI for predictions of glm_ft (Logit Model: Age, DTI, Income, Race, Sex)

```{r}
glm_ft_prediction <- 
  predict(glm_ft, mortgage_loan_glm, type = "response", se.fit = T)

summary(glm_ft_prediction$fit - glm_ft_prediction$se.fit)
summary(glm_ft_prediction$fit + glm_ft_prediction$se.fit)

```

Logit Model: Age, DTI, Income, Race, Sex, Loan Amount

```{r}
glm_ft1 <- 
  glm(loan_app_denied ~ 
        factor(applicant_age) + factor(debt_to_income_group) + income + loan_amount +factor(derived_race) + 
        factor(derived_sex) -1, 
      family = binomial(link = "logit"),
      data = mortgage_loan_glm)
summary(glm_ft1)

mortgage_loan_glm$glm_ft1_predict <- 
  ifelse(predict(glm_ft1, mortgage_loan_glm, type = "response") < 0.5, 0, 1)

table(mortgage_loan_glm[c("loan_app_denied","glm_ft1_predict")])
100*table(mortgage_loan_glm$glm_ft1_predict==mortgage_loan_glm$loan_app_denied)/nrow(mortgage_loan_glm)

save(glm_ft1, file = '/Users/thidathornvanitsthian/Documents/Data Science 2022/loan-denial-rate-TN/Data/glm_ft1')
```

Create CI for predictions of glm_ft1 (Logit Model: Age, DTI, Income, Race, Sex, Loan Amount)

```{r}
glm_ft1_prediction <- 
  predict(glm_ft1, mortgage_loan_glm, type = "response", se.fit = T)

summary(glm_ft1_prediction$fit - glm_ft_prediction$se.fit)
summary(glm_ft1_prediction$fit + glm_ft_prediction$se.fit)
```



Income is significant by itself

```{r}
glm_income <- 
  glm(loan_app_denied ~ income, 
      family = binomial(link = "logit"),
      data = mortgage_loan_glm)
summary(glm_income)
confint(glm_income)

glm_income_predict <- 
  predict(glm_income, mortgage_loan_glm)

mortgage_loan_glm$glm_income_predict <- 
  ifelse(predict(glm_income, mortgage_loan_glm, type = "response") < 0.5, 0, 1)


100*table(mortgage_loan_glm$glm_income_predict==mortgage_loan_glm$loan_app_denied)/nrow(mortgage_loan_glm)

```

```{r}
glm_race <- 
  glm(loan_app_denied ~ factor(derived_race), 
      family = binomial(link = "logit"),
      data = mortgage_loan_glm)
summary(glm_race)
confint(glm_race)

glm_race_predict <- 
  predict(glm_race, mortgage_loan_glm)

mortgage_loan_glm$glm_race_predict <- 
  ifelse(predict(glm_race, mortgage_loan_glm, type = "response") < 0.5, 0, 1)

save(glm_race, file = '/Users/thidathornvanitsthian/Documents/Data Science 2022/loan-denial-rate-TN/Data/glm_race')

100*table(mortgage_loan_glm$glm_race_predict==mortgage_loan_glm$loan_app_denied)/nrow(mortgage_loan_glm)
```

```{r}
glm_DTI <- 
  glm(loan_app_denied ~ factor(debt_to_income_group), 
      family = binomial(link = "logit"),
      data = mortgage_loan_glm)
summary(glm_DTI)
confint(glm_DTI)

glm_DTI_predict <- 
  predict(glm_DTI, mortgage_loan_glm)

mortgage_loan_glm$glm_DTI_predict <- 
  ifelse(predict(glm_DTI, mortgage_loan_glm, type = "response") < 0.5, 0, 1)


100*table(mortgage_loan_glm$glm_race_predict==mortgage_loan_glm$loan_app_denied)/nrow(mortgage_loan_glm)
```






